A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-03-10 analysis/paper/paper.Rmd
Using saa style guide. Trying also to include all dates in the comparison

---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo, Institute of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
---

@@

The method of shoreline dating has ranged from being met with scepticism as related to the degree sites would have been consistently shore-bound, to being characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer little more than a earliest possible date for when a site could have been in use. The most common application in Norway has arguably been to use the method to provide an approximate date for the occupation of the sites, typically in combination with other dating methods [see for example chapters in @; ; and below]. Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time, typically by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [e.g. @breivik2014; @breivik2018a; @solheim2018; @fossum2020; @nielsen2021b]. In his discussion of the method, @nordqvist1999, in part drawing on @johansen1963, argues that there can be little doubt about the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer.

@@

This process was repeated 1000 times for each phase for each site. The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5, available in the supplementary material [cf. @crema2010, 1125]. Hovland 5 was chosen for this evaluation as it has a fairly uncertain date range, and is located in area of quite complex topography. At 1000 simulation runs the analysis of each site took up to a few hours, depending on the distance to the simulated shorelines and the necessary size of the window of analysis.

@@

Where x is [...] In Figure \@ref(fig:shoredate) this formula is used to shoreline date the same sites from where this relationship was derived. Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some deviance in elevation over the site limits. The result of comparing the radiocarbon and shoreline dates using the method presented by @parnell2008 for assessing synchroneity is given in Figure \@ref(fig:shoredate)B. Here, 100,000 age samples drawn from each shoreline date was subtracted from 100,000 age samples drawn from the corresponding ^14^C-date. The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered to be in agreement. The result of this procedure indicates that the shoreline date correspond with the radiocarbon dates for 46 out of 49 sites (`r round(46/49, 2)*100`%). The deviation from the shoreline date at Gunnarsrød 5 is to be expected based on the simulation results (see above), but not at Langemyr and Kvastad A2. For these two sites the zero age difference can be seen to fall just outside the 95% interval (Figure \@ref(fig:shoredate)B). If one accepts that sensitivity in the method leads to the failure to accurately date these two sites, and that the failed dating of Gunnarsrød 5 is instead due to extraneous factors related the DTM and not the method itself, one could alternatively see the success rate as defined by 44 out of 46 sites (`r round(44/46, 2)*100`%) having been correctly dated. However, a different implication of the failed dating of these two sites could be that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for the earliest part of the Mesolithic (cf. Figure \@ref(fig:results)).

@@

To further explore the implementation for shoreline dating outlined above, excavated and shoreline-dated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites have been subjected to the method (Figure \@ref(fig:redate)). The resulting dates are compared to those originally proposed in the excavation reports for the sites. To avoid issues with recent disturbances in the DTM, the sites have been dated based on the mean of the altitudes provided in the report for each site. The results highlight the spatial and temporal contingency of the method, illustrated by the variation in the range of the 95% intervals for the dates. In some cases the method provides a very precise date range and in others it offers little more than a *terminus post quem*. This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise dates (Figure \@ref(fig:map-iso)B).

The comparison with previously reported dates is an illustrative, but unfair exercise for a few reasons. First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an implied or explicit, but undefined uncertainty range. Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change. Finally, they are often only meant to indicate a lower bound for when the sites could have been in use. Nonetheless, I do think that the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner [although not always, e.g. @roalkvam2022]. That being said, the results do indicate that shoreline dating has at times been applied with an exaggerated degree of precision, and while the implications of a more stable RSL-change for the precision of shoreline dating is known, it appears to be somewhat under-appreciated. Furthermore, some of the date ranges resulting from the method outlined here extend well beyond major chronological divisions, even into the Iron Age, and could be severely and securely constrained with only cursory reference to typology. This is arguably an exercise  This also points to the possibility of drawing on other temporal data, for example within a Bayesian framework, to improve the precision of the dates.

@@

Not least following from the fact that few Preboreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g. @bang-andersen2012; @breivik2014; @glorstad2016]. The shoreline dated pre-Boreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a]. These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision use of the sites after the sea retreated any significant distance from their location. In the original publication of the sites, @jaksland2014 provides a thorough discussion of shoreline dating in general, and as used for the dating of the Brunlanes sites specifically. A comparison of his results and the ones achieved using the above-outlined approach are given in Figure \@ref(fig:brunlanes)A. The small discrepancies between the results mainly follow from the fact that a slightly updated version of the displacement curve is applied here [cf. @sorenseninpress]. Jaksland's dates are given a flat 200 and 50 year uncertainty range starting from what he gives as the earliest possible date. The 200 year uncertainty range is given if the sites were to be considered in isolation, while the argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other. Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they don't overlap. This information is not integrated in the approach outlined here, but could justify reducing the the uncertainty ranges. Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does nonetheless result in very precise dates. Above it was shown how additional temporal data could be combined with the method to improve its accuracy and precision. This example, on the other hand, highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use. One approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the steepness of the slope to the shoreline. Based on this, it could be possible to exclude certain elevations as unlikely for the position of the shoreline for when the site was in use. Such approaches would make less of an impact in this setting, where the 95% HDR is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change has been less severe (cf. Figure \@ref(fig:redate)).

@@

It is worth highlighting some sources of likely variation and uncertainty that have not been considered here. First of all the DTM has only been corrected for major modern disturbances. This means that erosion, although likely not that prevalent, has not been taken into account. Secondly, the DTM has a vertical error [cf. @lewis2021]. Thirdly, the displacement curves were here interpolated to all site locations without accounting for increased uncertainty as one moves away from the isobases of the displacement curves. Finally, the definition of the site limits nor the elevation range over which these extend was not considered.

The most immediate contribution of this paper is what must be considered a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline. This is indicated by the close relationship between sites and the shoreline up until the transition to the Neolithic at c. 3900 BCE, after which a couple of sites become situated some distance from the shoreline. This is followed by a more decisive break at the transition to the Late Neolithic at c. 2400 BCE -- in clear agreement with the literature. Furthermore, based on the quantitative nature of these findings, a refined method for the shoreline dating of Stone Age sites has been proposed. This involves both taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, and implementing formula x to account for the distance between the sites and the shoreline. When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratios identified in Figure \@ref(fig:redate) to characterise this relationship. However, the accuracy of the method can be improved by including more information, both with reference to the topographic location of the sites and other temporal data such as radiocarbon dates and typological indicators in the artefact inventories. The precision of the method is, as shown above, both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more precise date than younger sites located towards the south-west. The impact of such additional information will therefore also vary.

Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further evaluate and adjust the findings reported here, but a larger sample size could also lay the foundations for refining the method by identifying subsets of sites for which the application of the method could be adjusted and refined. Given it's behavioural nature, it would for example seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline they were located. Furthermore, other dimensions related to the topographic location of the sites could be similarly explored. This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [e.g. @roalkvam2020], and which presumably has implications for how close to the shoreline people settled [helskog1978; @blankholm2020]. This is also related to the fact that while the mean sea-level is used for dating the sites, a consideration of the tidal range could possibly also have implications for the site location relative to the shoreline, depending on the topography [@helskog1978]. If such locational patterns can be discerned, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the nature of the occupation of the site beyond its location is available.

Finally, this analysis was based on a simulation approach for integrating multiple sources of spatio-temporal uncertainty. Here this was simply used to inform the question of the distance between sites and shorelines. However, this method can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.

This approach will be useful either when shoreline dating a large number of sites where no other information than site elevation is taken into account , which is typically done using a uniform range of uncertainty, likely overestimating the precision of the method in most cases and possibly underestimating it in some.

>>>>

---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \usepackage{fancyhdr}
   - \linenumbers
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo, Institute of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
---

<!--- Page numbers in margin to avoid conflict with figure captions --->
\fancypagestyle{plain}{%
  \fancyhf{}
  \fancyfoot[R]{\leavevmode\rlap{\thepage~}}
}
\pagestyle{plain}

@@

The method of shoreline dating has been met with scepticism as related to the degree sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than a earliest possible date for when a site could have been in use [@breivik2018a, 174]. The most common application in Norway has arguably been to use the method to provide an approximate date for the occupation of the sites, typically in combination with other dating methods [see for example chapters in @; ; and below]. Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time, typically by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [e.g. @breivik2014; @breivik2018a; @solheim2018; @fossum2020; @nielsen2021b]. In his discussion of the method, @nordqvist1999, in part drawing on @johansen1963, argues that there can be little doubt concerning its general applicability -- what is less clear is the level of reliability and chronological resolution that the method can offer.

@@

This process was repeated 1000 times for each phase for each site. The choice of 1000 simulation runs follows from an evaluation of when the mean distances between site and shoreline converged when running 5000 iterations of the simulation on the site Hovland 5, available in the supplementary material [cf. @crema2010, 1125]. Hovland 5 was chosen for this evaluation as it has a fairly uncertain date, and is located in area of quite complex topography.

@@

Where x is [...] In Figure \@ref(fig:shoredate) this formula is used to shoreline date the same sites from where this relationship was derived. Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some deviance in elevation over the site limits. The result of comparing the radiocarbon and shoreline dates using the method presented by @parnell2008 for assessing the synchroneity between dates is given in Figure \@ref(fig:shoredate)B. Here, 100,000 age samples drawn from the probability density function of each shoreline date was subtracted from 100,000 age samples drawn from the corresponding ^14^C-date. The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered to be in agreement. The result of this procedure indicates that the shoreline date correspond to the radiocarbon dates for 46 out of 49 sites (`r round(46/49, 2)*100`%). The deviation from the shoreline date at Gunnarsrød 5 is to be expected based on the simulation results (see above), but not at Langemyr and Kvastad A2. For these two sites the zero age difference can be seen to fall just outside the 95% interval (Figure \@ref(fig:shoredate)B). If one accepts that sensitivity in the method leads to the failure to accurately date these two sites, and that the failed dating of Gunnarsrød 5 is instead due to extraneous factors related the DTM and not the method itself, one could alternatively see the success rate as defined by 44 out of 46 sites (`r round(44/46, 2)*100`%) having been correctly dated. However, another possible implication of the failed dating of these two sites could be that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for sites from the earliest part of the Mesolithic (cf. Figure \@ref(fig:results)).

@@

To further explore the implementation for shoreline dating outlined above, excavated and shoreline-dated Stone Age sites within the study area where ^14^C-dates are not available or are not believed to date the main occupation of the sites have been subjected to the method (Figure \@ref(fig:redate)). The resulting dates are compared to those originally proposed in the excavation reports for the sites (the numerical results are available in the supplementary material). To avoid issues with recent disturbances on the DTM, the sites have been dated based on the mean of the altitudes provided in the report for each site. The results highlight the spatial and temporal contingency of the method, illustrated by the variation in the range of the 95% intervals for the dates. In some cases the method provides a very precise date range and in others it offers little more than a *terminus post quem*. This is dependent on the steepness of the displacement curves, leading to the general pattern of older sites situated towards the north-east getting more precise dates (cf. Figure \@ref(fig:map-iso)B).

The comparison with previously reported dates is an illustrative, but unfair exercise for a few reasons. First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an implied or explicit, but undefined uncertainty range. Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change. Finally, they are sometimes only meant to indicate a lower bound for when the sites could have been in use. Nonetheless, the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner [although see e.g. @roalkvam2022]. That being said, the results do also indicate that shoreline dating has at times been applied with an exaggerated degree of precision, and while the implications of a more stable RSL-change for the duration and re-use of site locations, and consequently the precision of shoreline dating is well known, it appears to be somewhat under-appreciated. Some of the date ranges resulting from the method outlined here clearly extend well beyond major chronological divisions, even into the Iron Age, and could be severely and securely constrained with only cursory reference to typology. However, while this is obvious in some cases, the nature and uncertainty inherent to the method still means that this is arguably a required exercise that should be made explicit. This final point also points to the possibility of drawing on other temporal data, for example within a Bayesian framework, to further improve the precision of the dates that can be achieved with shoreline dating.

@@

Not least following from the fact that few Preboreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g. @bang-andersen2012; @breivik2014; @glorstad2016]. The shoreline-dated Preboreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a]. These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision use of the sites after the sea retreated any significant distance from their location. In the original publication of the sites, @jaksland2014 provides a thorough discussion of shoreline dating in general, and as used for the dating of the Brunlanes sites specifically. A comparison of his results and the ones achieved using the above-outlined approach are given in Figure \@ref(fig:brunlanes)A. The sites have been dated using what @jaksland2014 gives as the lowest elevation of finds at each site, and by employing a exponential decay ratio of , to allow for more deviance in the distance between site and shoreline. This corresponds to the decay ratio for sites older than 7000 BCE in Figure \@ref(fig:results2).

The small discrepancies between the achieved results mainly follow from the fact that a slightly updated version of the local displacement curve is applied here [cf. @sorenseninpress]. Jaksland's dates are given a flat 200 and 50 year uncertainty range starting from what he gives as the earliest possible date. The 200 year uncertainty range is given if the sites were to be considered in isolation, while the argument for the uncertainty range of only 50 years is based on the location of the sites relative to each other. Since they are located in such a constrained and steep area of the landscape, the difference in elevation between the sites is argued to establish their relative date and thus constrain the uncertainty ranges so that they don't overlap. This information is not integrated in the approach outlined here, but could justify further reducing the uncertainty ranges. Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does nonetheless result in very precise dates. Above it was shown how additional temporal data could be combined with the method to improve its accuracy and precision. This example, on the other hand, highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help in increasing the precision of the method if this can be used to exclude certain sea-levels as unlikely for when a site was in use. One approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the distance and steepness of the slope to the shoreline. Based on this, it could conceivably be possible to exclude certain elevations as unlikely for the position of the shoreline when the site was in use. Such approaches would make less of an impact in this setting, where the 95% HDR is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change has been less severe (cf. Figure \@ref(fig:redate)).

@@

It is worth highlighting some sources of likely variation and uncertainty that have not been considered here. First of all the DTM has only been corrected for major modern disturbances. This means that erosion, although likely not that prevalent, has not been taken into account. Secondly, the DTM has a vertical error [cf. @lewis2021]. Thirdly, the displacement curves were here interpolated to all site locations without accounting for increased uncertainty as one moves futher away from the isobases of the displacement curves. Finally, the question of how site limits are defined nor the elevation range over which these extend was given much consideration. While each of these factors will have variable but likely not too large of an impact on the results, they clearly represent dimensions which would all benefit from further exploration.

The most immediate contribution of this paper is what must be considered a confirmation of previous research into the relation between coastal Norwegian Stone Age sites and the prehistoric shoreline. This is indicated by the close relationship between sites and the shoreline up until the transition to the Neolithic at c. 3900 BCE, after which a couple of sites become situated some distance from the shoreline. This is followed by a more decisive break at the transition to the Late Neolithic at c. 2400 BCE -- in clear agreement with the literature. Furthermore, based on the quantitative nature of these findings, a refined method for the shoreline dating of pre-Late Neolithic Stone Age sites has been proposed. This involves both taking the distance between sites and the isobases of the displacement curves into consideration when dating the sites, and implementing formula X to account for the distance between the sites and the shoreline. When no other information is available, it can at present be recommended to use the empirically derived exponential decay ratio identified in Figure \@ref(fig:redate) to characterise this relationship. However, the accuracy of the method can be improved by including more information, both with reference to the topographic location of the sites and other temporal data such as radiocarbon dates and typological indicators in the artefact inventories. The precision of the method is, as shown above, both geographically and temporally contingent due to the trajectory of RSL-change, where older sites situated towards the north-east in the study area will get a more precise date than younger sites located towards the south-west. The impact of such additional information will therefore also vary.

Future investigations and radiocarbon dates from Stone Age sites in the region can not only be used to further evaluate and adjust the findings reported here, but a larger sample size could also lay the foundations for refining the method by identifying subsets of sites for which the application of the method could be adjusted and refined. Given it's behavioural nature, it would for example seem likely that dimensions such as the nature and purpose of visits to the sites will have implications for how close to the shoreline they were located. Furthermore, other dimensions related to the topographic location of the sites could be similarly explored. This for example pertains to the exposure of sites to wave action, which is likely to have been of concern [e.g. @roalkvam2020], and which presumably has implications for how close to the shoreline people settled [@helskog1978; @blankholm2020]. This is also related to the fact that while the mean sea-level is used for dating the sites, a consideration of the tidal range could possibly also have implications for the site location relative to the shoreline, depending on the topography [@helskog1978]. Dimensions such as these was given a cursory treatment here, with the estimation of the horizontal and topographic distance to the shoreline, but these dimensions could also be investigated further. If patterns related to such locational patterns can be discerned, this will not least be useful for improving the shoreline dating of sites which have only been surveyed and where little information on the site beyond its location is available.

Finally, this analysis was based on a simulation approach for integrating multiple sources of spatio-temporal uncertainty. Here this was simply used to inform the question of the distance between sites and the shore-line. However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.
